---
title: "MovieLens"
author: "Mauro Berlanda"
date: "4/5/2020"
output: pdf_document
sansfont: Calibri Light
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- http://rmarkdown.rstudio.com -->

## Introduction/Overview/Executive Summary

<!-- an introduction/overview/executive summary section that describes the dataset and summarizes the goal of the project and key steps that were performed -->

The goal of this project is to develop an algorithm to predict movie ratings.

This analysis is performed on ([MovieLens](https://grouplens.org/datasets/movielens/10m/)) dataset. It contains 10 millions ratings and 100 000 tags applications applied to 10 000 movies by 72 000 users.

```{r data, include=FALSE}
# The following lines assume that you run the MovieLens.R script before
# compiling the markdown. Dumping some intermediary data structures will allow
# to make the md compilation lighter

require(tidyverse)
require(caret)
require(data.table)
load("movie-data.rda")
```

After downloading the entire dataset, two dataframes are created:

* `edx` which is used to perform the analysis and develop the algorithm to predict ratings
* `validation` which contains the true values of the predictions


```{r edx, echo=TRUE}
dim(edx)
dim(validation)
colnames(edx)
```

The key steps required to identify an optimal solution are:

  * todo
  * create a train and a validation set
  * explore the content of the train set
  * another step


## Method/Analysis

<!--  a methods/analysis section that explains the process and techniques used, including data cleaning, data exploration and visualization, insights gained, and your modeling approach -->

```{r ratings}
boxplot(edx$rating)
title("Ratings distribution in edx train set")
summary(edx$rating)
```

## Results

<!-- a results section that presents the modeling results and discusses the model performance -->

## Conclusion

<!-- a conclusion section that gives a brief summary of the report, its limitations and future work -->
