---
title: "Harvard PH125.9x Capstone: Mushroom Classification Analysis"
author: "Mauro Berlanda"
date: "April 2020"
output: pdf_document
sansfont: Calibri Light
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo=TRUE,
  warning=FALSE,
  cache=FALSE,
  fig.align='center'
)
```

## Introduction

The last assignment of the [Datascience Professional Certificate](https://courses.edx.org/dashboard/programs/3c32e3e0-b6fe-4ee4-bd4f-210c6339e074/) by HarvardX on edx is submitting its own report.
The main goal of the project is to prove the ability to clearly communicate the process and the insights gained from an analysis.

We are going to use for this analysis the [Mushroom records drawn from The Audubon Society Field Guide to North American Mushrooms (1981](http://archive.ics.uci.edu/ml/datasets/Mushroom). This data set includes descriptions of hypothetical samples corresponding to 23 species of gilled mushrooms in the Agaricus and Lepiota Family.

> The csv file containing the data was originally downloaded from [Kaggle](https://www.kaggle.com/uciml/mushroom-classification/data) due to its ease of manipulation. The file has been commited in a github repository since Kaggle downloads require authentication. Being unable to retrieve the raw zip file due to a corrupted output (`unzip error -1`), my script is downloading the uncompressed csv file. It does not exceed 365Kb, so it can be requested without any performance or network traffic concern.

```{r load_deps, include=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(matrixStats)) install.packages("matrixStats", repos = "http://cran.r-project.org")

if(!require(e1071)) install.packages("matrixStats", repos = "http://cran.r-project.org")
if(!require(gam)) install.packages("gam", repos = "http://cran.r-project.org")
```

```{r data_download_parse}
file_url <- "https://raw.githubusercontent.com/mberlanda/ph125-9x-data-science-capstone/master/mushroom-classification/mushrooms.csv"
csv_filepath <- "./mushrooms.csv"

# Download the csv file if needed
if (!file.exists(csv_filepath)) {
  download.file(file_url, csv_filepath)
}

# Use read.csv to parse the file converting strings to factors
mushrooms <- read.csv(csv_filepath, header=TRUE, sep=",", stringsAsFactors=TRUE)
# Explore the columns and types of the dataset
str(mushrooms)

rm(file_url, csv_filepath)
```

To improve the domain knowledge, you can find below an image illustrating the different [parts of a mushroom](https://ohioline.osu.edu/factsheet/plpath-gen-11):

```{r mushrooms_img, echo=FALSE, out.width='100%'}
mushroom_img <- './HYG_3303_mushroom_parts.jpg'
mushroom_img_url <- 'https://ohioline.osu.edu/sites/ohioline/files/imce/Plant_Pathology/HYG_3303_mushroom_parts.jpg'
if (!file.exists(mushroom_img)){
    download.file(mushroom_img, mushroom_img)
}
knitr::include_graphics(mushroom_img)
rm(mushroom_img, mushroom_img_url)
```

All the attributes are factors and they represent the following abbreviations:

1. cap-shape: bell=b,conical=c,convex=x,flat=f, knobbed=k,sunken=s
2. cap-surface: fibrous=f,grooves=g,scaly=y,smooth=s
3. cap-color: brown=n,buff=b,cinnamon=c,gray=g,green=r, pink=p,purple=u,red=e,white=w,yellow=y
4. bruises?: bruises=t,no=f
5. odor: almond=a,anise=l,creosote=c,fishy=y,foul=f, musty=m,none=n,pungent=p,spicy=s
6. gill-attachment: attached=a,descending=d,free=f,notched=n
7. gill-spacing: close=c,crowded=w,distant=d
8. gill-size: broad=b,narrow=n
9. gill-color: black=k,brown=n,buff=b,chocolate=h,gray=g, green=r,orange=o,pink=p,purple=u,red=e, white=w,yellow=y
10. stalk-shape: enlarging=e,tapering=t
11. stalk-root: bulbous=b,club=c,cup=u,equal=e, rhizomorphs=z,rooted=r,missing=?
12. stalk-surface-above-ring: fibrous=f,scaly=y,silky=k,smooth=s
13. stalk-surface-below-ring: fibrous=f,scaly=y,silky=k,smooth=s
14. stalk-color-above-ring: brown=n,buff=b,cinnamon=c,gray=g,orange=o, pink=p,red=e,white=w,yellow=y
15. stalk-color-below-ring: brown=n,buff=b,cinnamon=c,gray=g,orange=o, pink=p,red=e,white=w,yellow=y
16. veil-type: partial=p,universal=u
17. veil-color: brown=n,orange=o,white=w,yellow=y
18. ring-number: none=n,one=o,two=t
19. ring-type: cobwebby=c,evanescent=e,flaring=f,large=l, none=n,pendant=p,sheathing=s,zone=z
20. spore-print-color: black=k,brown=n,buff=b,chocolate=h,green=r, orange=o,purple=u,white=w,yellow=y
21. population: abundant=a,clustered=c,numerous=n, scattered=s,several=v,solitary=y
22. habitat: grasses=g,leaves=l,meadows=m,paths=p, urban=u,waste=w,woods=d

The classes used for the outcome are `edible` or `poisonous`. In the Kaggle version of the data set there is no `unknown` classification value and missing values have been removed:

```{r missing_values}
unique(mushrooms$class)
sum(is.na(mushrooms))
```
